BACI of biomass change for VOCC source cells

### 2019-08-23 analysis/VOCC/4-vocc-BACI.Rmd
refine matchit: best seems to be combined exact matching of density and nearestneighbor for other variables

---
title: "BACI of biomass change for VOCC source cells"
author: "Philina English"
date: '2019-08-15'
output: html_document
---

@@

#### Set species
```{r echo=TRUE}
# # Species run so far...
# species <- "Arrowtooth Flounder"
# species <- "Pacific Cod"
# species <- "Sablefish"
species <- "Silvergray Rockfish"
# species <- "Lingcod"
# species <- "North Pacific Spiny Dogfish" # note: using all data for maturity thresholds
# species <- "Quillback Rockfish"
# species <- "Pacific Ocean Perch"
# species <- "Yelloweye Rockfish"
```

@@

#### Make climate vectors for biannual changes between surveys
```{r echo=TRUE}
trend <- FALSE
start_time <- 2012
end_time <- 2015
temp_threshold <- c(0.5)
do_threshold <- c(0.1)
```

@@

Plot all possible vectors
```{r echo=FALSE}
if (climate == "temperature") {
  vocc$diff <- vocc$units_per_decade / 10 * delta_t_total

@@

Filter vectors to match optimal range for chosen species
```{r}
vectors <- trim_vector_data(vocc, variable_names, optimal_values, min_thresholds, max_thresholds,
  cell_size = input_cell_size, dist_intercept = input_cell_size/2,
  max_dist = 120
)

@@

### Plot remaining climate vectors for regions where change exceeds thresholds
```{r echo=FALSE, warning=FALSE}

if (climate == "temperature") {
  vectors$diff <- vectors$units_per_decade / 10 * delta_t_total

@@

#### Get all biomass predictions 
```{r, echo=FALSE, warning=FALSE}
priors <- FALSE
# priors <- TRUE

@@

#### Extract biomass values to match spatial and temporal range of climate vectors

```{r}
#i <- biomass_list[[1]]

vocc_by_sp <- lapply(biomass_list, function(i) {
  biomass_change <- make_trend_data(i,
    ssid = model_ssid,
    start_time = start_time,
    skip_time = skip_time,
    end_time = end_time,
    input_cell_size = input_cell_size,
    scale_fac = scale_fac,
    delta_t_total = delta_t_total,
    delta_t_step = delta_t_step,
    indices = indices,
    variable_names = "est_exp"
  )

@@

#browser()
  alldata <- rbind(source_biomass, control_biomass) 
  alldata <- mutate(alldata, treat = ifelse(cell_type=="source", 1, 0)) %>% 
    select(-var_2_min, -var_2_max) %>% 
    rename(vect_dist = distance)
  alldata$treat <- as.factor(alldata$treat)
  ggplot(alldata, aes(x=cell_type, y=log(start_density))) + geom_boxplot()

  ## still including this trimming to reduce size of dataset 
  control_min <- min(alldata$start_density[alldata$treat=="0"])
  source_min <- min(alldata$start_density[alldata$treat=="1"])
  # control_min <- quantile(alldata$start_density[alldata$treat=="0"], 0.001)
  # source_min <- quantile(alldata$start_density[alldata$treat=="1"], 0.001)
  group_min <- max(control_min, source_min)

  control_max <- max(alldata$start_density[alldata$treat=="0"])
  source_max <- max(alldata$start_density[alldata$treat=="1"])
  # control_max <- quantile(alldata$start_density[alldata$treat=="0"], 0.95)
  # source_max <- quantile(alldata$start_density[alldata$treat=="1"], 0.95)
  group_max <- min(control_max, source_max)
  # 
   alldata <- filter(alldata, start_density >= group_min) #%>% filter(start_density <= group_max)
  # 
  # num_source <- nrow(alldata[alldata$treat=="1", ])
  #

  # alldata <- filter(alldata, start_density >= control_min)

  # low_biomass <- quantile(alldata$start_density, 0.1)
  # alldata <- filter(alldata, start_density >= low_biomass )

  # if (order == "decreasing") 
  #alldata <-   alldata[rev(order(alldata$start_density)), ]
  # 
  # if (order == "increasing") 
 # alldata <- alldata[rev(order(alldata$start_density)), ]

@@

 ## Using mysterious cem optimization ... doesn't work   
  # fit_formula <- paste0("treat ~ start_density_st") 
  # fit_formula <- paste0("treat ~ start_density_st + depth.st") 
  # fit_formula <- paste0("treat ~ start_density_st + depth + trawled + muddy + any_rock + omega_s + x + y") 
  # fit_formula <- paste0("treat ~ start_density_st + depth.st + muddy.st + any_rock.st") #+ + omega_s
  # 
  # # for (j in (variable_names)) {
  # #   fit_formula <- paste0("", fit_formula, " + ", j, "_s")
  # #   }
  # 
  # # # options("optmatch_max_problem_size" = Inf)
  # # matched <- MatchIt::matchit(as.formula(fit_formula), data = alldata, method = "optimal", ratio = 1)
  # #fit_formula <- paste0("treat ~ depth.st")
  # matched <- MatchIt::matchit(as.formula(fit_formula),
  #   method = "cem",
  #   distance = "probit",
  #   discard="control",
  #   data = alldata)

  # Using exact match and rounding
  fit_formula <- paste0("treat ~ depth.st +  muddy.st + any_rock.st") #+ trawled + omega_s
  for (j in (variable_names)) {
    fit_formula <- paste0("", fit_formula, " + ", j, "_s")
    }

  matched <- MatchIt::matchit(as.formula(fit_formula), exact = c("start_density.st"),
    method = "nearest", m.order = "random",
    #method = "full",
    distance = "probit",
    discard="control",
    data = alldata)

  mdata <- MatchIt::match.data(matched, distance = "pscore")

  summary(matched)
  plot(matched)

@@

  # controls2 <- mdata[matched$match.matrix, ]
  # controls2$origobs <- row.names(controls)   
  # controls2$matchobs <- matched$match.matrix[,2]

@@

  #mdata <- mdata %>% rename(distance = vect_dist) #%>% select(-start_density.st, -depth.st, -muddy.st, -any_rock.st, -omega_s.st)

@@

  # rename spatial coordinates for inclusion in models and plots without naming conflicts
  vocc_by_sp <- vocc_by_sp %>%
    rename(X = x, Y = y)

  vocc_by_sp
})
```

## Adults

Prep data
```{r}
data <- vocc_by_sp [["adult"]]
# View(data)

data$log_density <-  log(data$density)
data$vect_dist2 <- data$vect_dist^2

data <- mutate(data, time = ifelse(after==1, "end", "begin"))
data$matchobs <- as.factor(data$matchobs)
data$origobs  <- as.factor(data$origobs)   
```

Plot data
```{r}
ggplot(data, 
#ggplot(filter(data, as.numeric(matchobs) <= 550), 
  aes(y=Y, x=X, shape=as.factor(treat), colour=matchobs)) +
  geom_point(alpha=0.4, size =4) +
  scale_shape_manual(values= c(4, 1)) +
  scale_colour_viridis_d(option="C", guide = FALSE) +
  theme_bw()

@@

Basic BACI
```{r}
for (j in (variable_names)) {
  baci_formula <- paste0("log_density~after + cell_type + after:cell_type + n_targets + vect_dist + vect_dist2 + std_", j, "+ std_", j, "2 + (1|icell) + (1|matchobs)")
}

@@

GAMM
```{r}

for (j in (variable_names)) {
  gamm_formula <- paste0("log_density ~ time * cell_type + s(vect_dist) + s(std_", j, ") + s(X,Y)")
}

@@

# for (j in (variable_names)) {
#   gamm_formula <- paste0("log_density ~ time * cell_type + s(vect_dist) + s(std_", j, ") + te(X,Y)")
# }
# gam <- mgcv::gamm(as.formula(gamm_formula), random = ~(1|matchobs) + (1|origobs), data=data) # + s(X,Y, bs="re")

plot(gam4$gam,pages=1)

summary(gam4$mer) ## underlying mixed model
anova(gam4$gam)

# itsadug::plot_parametric(gam4$gam, pred=list(cell_type=c("control", "source"), after=c("0","1")), 
#   parametricOnly = TRUE, main="") + scale_y_reverse()

plot_parametric(gam4$gam, pred=list(time=c("begin","end"), cell_type=c("control", "source")), 
   parametricOnly = TRUE, main="", xlab ="log_density")

summary(gam4$gam) ## summary of gam
```

## Immatures

Prep data
```{r}
idata <- vocc_by_sp[["imm"]]
#View(data)
idata$log_density <-  log(idata$density)
idata$vect_dist2 <- idata$vect_dist^2

idata <- mutate(idata, time = ifelse(after==1, "end", "begin"))
idata$matchobs <- as.factor(idata$matchobs)
idata$origobs  <- as.factor(idata$origobs)   
```

Plot data
```{r}
ggplot(filter(data, as.numeric(matchobs) <= 150), aes(y=Y, x=X, shape=as.factor(treat), colour=matchobs)) +
  geom_point(alpha=0.4, size =4) +
  scale_shape_manual(values= c(4, 1)) +
  scale_colour_viridis_d(option="C", guide = FALSE) +
  theme_bw()

@@

Basic BACI
```{r}
mi <- lmerTest::lmer(as.formula(baci_formula), data=idata)
summary(mi)
```

```{r}
sjPlot::plot_model(mi, type = "std2", title = paste(species, "immature"))
```

GAMM
```{r}
# for (j in (variable_names)) {
#   gamm_formula <- paste0("log_density ~ time * cell_type + s(vect_dist) + s(std_", j, ") + s(X,Y)")
# }

@@

plot(igam4$gam,pages=1)

summary(igam4$gam) ## summary of gam
summary(igam4$mer) ## underlying mixed model
anova(igam4$gam)

# itsadug::plot_parametric(gam4$gam, pred=list(cell_type=c("control", "source"), after=c("0","1")), 
#   parametricOnly = TRUE, main="") + scale_y_reverse()

#my_cols <- c("#E69F00", "#56B4E9")
plot_parametric(igam4$gam, pred=list(time=c("begin","end"), cell_type=c("control", "source")), 
   parametricOnly = TRUE, main="", xlab ="log_density"
  #, groups = cell_type, gcolor = my_cols, color = my_cols["time"]
  ) 
```

Misc exploring gamm 
```{r eval=FALSE}
# mi <- mgcv::gamm(log_density~ as.factor(after) + cell_type + after:cell_type + std_temp + std_temp2 + vect_dist + vect_dist2 + s(as.factor(idata$icell), bs="re") + s(as.factor(idata$matchobs), bs="re"), data=idata)
# summary(mi)
# 
# mi <- mgcv::gamm(log_density~ as.factor(after) * cell_type + s(std_temp) + s(vect_dist) + s(as.factor(idata$matchobs), bs="re") , data=idata) # + s(X,Y, bs="re")
# summary(mi)
# mi
# 
# 
# idata$after <- as.factor(idata$after)
# idata$matchobs <- as.factor(idata$matchobs)
# 
# m4b <- gamm4::gamm4(log_density~ after * cell_type + s(std_temp) + s(vect_dist) + #s(X) + s(Y) + 
#     s(X,Y), random = ~(1|matchobs), data=idata) # + s(X,Y, bs="re")
# 
# 
# plot(m4b$gam,pages=1)
# 
# summary(m4b$gam) ## summary of gam
# summary(m4b$mer) ## underlying mixed model
# anova(m4b$gam) 
# 
# gamtabs(m1)
# 
# #library(itsadug)
# plot_parametric(m4b$gam, pred=list(cell_type=c("control", "source"), after=c("0","1")))
# 
# library(mgcv)
```

>>>>

---
title: "BACI of biomass change for VOCC source cells"
author: "Philina English"
date: '2019-08-22'
output: html_document
---

@@

#### Set species
```{r echo=TRUE}
# # Species run so far...
 species <- "Arrowtooth Flounder"
# species <- "Pacific Cod"
# species <- "Sablefish"
# species <- "Silvergray Rockfish"
# species <- "Lingcod"
# species <- "North Pacific Spiny Dogfish" # note: using all data for maturity thresholds
# species <- "Quillback Rockfish"
# species <- "Pacific Ocean Perch"
# species <- "Yelloweye Rockfish"
```

@@

#### Make climate vectors for biannual changes between surveys
```{r echo=TRUE}
trend <- FALSE
start_time <- 2012
end_time <- 2015
temp_threshold <- c(1)
do_threshold <- c(0.1)
```

@@

### Plot all possible vectors
```{r echo=FALSE}
if (climate == "temperature") {
  vocc$diff <- vocc$units_per_decade / 10 * delta_t_total

@@

### Filter vectors to match optimal range for chosen species
```{r}
vectors <- trim_vector_data(vocc, variable_names, optimal_values, min_thresholds, max_thresholds,
  cell_size = input_cell_size, dist_intercept = input_cell_size/2,
  max_dist = 120
)

@@

### Plot remaining climate vectors for regions where change exceeds thresholds
```{r echo=FALSE, warning=FALSE}
if (climate == "temperature") {
  vectors$diff <- vectors$units_per_decade / 10 * delta_t_total

@@

Get all biomass predictions 
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=F}
priors <- FALSE
# priors <- TRUE

@@

Extract biomass values to match spatial and temporal range of climate vectors
```{r}
alldata <- lapply(biomass_list, function(i) {

  biomass_change <- make_trend_data(i,
    ssid = model_ssid,
    start_time = start_time,
    skip_time = skip_time,
    end_time = end_time,
    input_cell_size = input_cell_size,
    scale_fac = scale_fac,
    delta_t_total = delta_t_total,
    delta_t_step = delta_t_step,
    indices = indices,
    variable_names = "est_exp"
  )

@@

  alldata <- rbind(source_biomass, control_biomass) 
  alldata <- mutate(alldata, treat = ifelse(cell_type=="source", 1, 0)) %>% 
    select(-var_2_min, -var_2_max) %>% 
    rename(vect_dist = distance)
  alldata$treat <- as.factor(alldata$treat)

  alldata

})
```

```{r, include=FALSE, echo=FALSE}
lapply(alldata, function(i){ggplot(i, aes(x=cell_type, y=log(start_density))) + geom_boxplot()})
```

# Select matched control cells
```{r}
matched <- lapply(alldata, function(alldata) {  
  ## trim non-overlapping values to reduce size of dataset 
  control_min <- quantile(alldata$start_density[alldata$treat=="0"], 0.001)
  source_min <- quantile(alldata$start_density[alldata$treat=="1"], 0.001)
  # control_max <- quantile(alldata$start_density[alldata$treat=="0"], 1)
  # source_max <- quantile(alldata$start_density[alldata$treat=="1"], 1)

  group_min <- max(control_min, source_min)
  # group_max <- min(control_max, source_max)
  alldata <- filter(alldata, start_density >= group_min) #%>% filter(start_density <= group_max)

@@

 ## Using optimization ... doesn't work   
  # if(
  #   class(
  #   try(
  #     fit_formula <- paste0("treat ~ start_density_st + depth.st + muddy.st + any_rock.st") 
  #   # fit_formula <- paste0("treat ~ start_density_st") 
  #   # fit_formula <- paste0("treat ~ start_density_st + depth.st") 
  #   # # for (j in (variable_names)) {
  #   # #   fit_formula <- paste0("", fit_formula, " + ", j, "_s")
  #   # #   }
  #     # options("optmatch_max_problem_size" = Inf)
  #     matched <- MatchIt::matchit(as.formula(fit_formula), method = "optimal", data = alldata)
  #     , silent = TRUE)
  #     ) == "try-error"){

  # Using exact match and rounding
  fit_formula <- paste0("treat ~ depth.st + muddy.st + any_rock.st + trawled + omega_s + x + y") #
  # for (j in (variable_names)) {
  #   fit_formula <- paste0("", fit_formula, " + ", j, "_s")
  #   }

  # Choose data order
  # alldata <- alldata[(order(alldata$y)), ]
  # if (order == "increasing") alldata <- alldata[rev(order(alldata$start_density)), ]
  # if (order == "decreasing") alldata <- alldata[rev(order(alldata$start_density)), ]

  matched <- MatchIt::matchit(as.formula(fit_formula), exact = c("start_density.st"),
    method = "nearest", m.order = "random",
    distance = "probit",
    discard="control",
    data = alldata)

  #}

  matched

})
```

Print diagnostics for matching
```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
lapply(matched, function(i){  
  summary(i)
  plot(i)
  })
```

Prep matched data for analysis
```{r}
vocc_by_sp <- lapply(matched, function(matched) {    
  mdata <- MatchIt::match.data(matched, distance = "pscore")

@@

  # if selecting 2 matches
  # controls2 <- mdata[matched$match.matrix, ]
  # controls2$origobs <- row.names(controls)   
  # controls2$matchobs <- matched$match.matrix[,2]

@@

@@

  # rename spatial coordinates for inclusion in models and plots without naming conflicts
  data <- vocc_by_sp %>%
    rename(X = x, Y = y)
      #%>% rename(distance = vect_dist) %>% select(-start_density.st, -depth.st, -muddy.st, -any_rock.st, -omega_s.st)
  # other variable prep
  data$log_density <-  log(data$density)
  data$vect_dist2 <- data$vect_dist^2
  data <- mutate(data, time = ifelse(after==1, "end", "begin"))
  data$matchobs <- as.factor(data$matchobs)
  data$origobs  <- as.factor(data$origobs)

  data

})
```

## Adults
```{r}
data <- vocc_by_sp [["adult"]]
# View(data)
```

### Plot matches
```{r}
ggplot(data, 
#ggplot(filter(data, as.numeric(matchobs) <= 550), 
  aes(y=Y, x=X, shape=as.factor(treat), colour=as.factor(matchobs))) +
  geom_point(alpha=0.4, size = 2) +
  scale_shape_manual(values= c(0, 15)) +
  scale_colour_viridis_d(option="C", guide = FALSE) +
  theme_bw()

@@

### Basic BACI
```{r}
for (j in (variable_names)) {
  baci_formula <- paste0("log_density~after + cell_type + after:cell_type + n_targets + vect_dist + vect_dist2 + std_", j, "+ std_", j, "2 + (1|icell) + (1|matchobs)")
}

@@

### GAMM
```{r}
for (j in (variable_names)) {
  gamm_formula <- paste0("log_density ~ time * cell_type + s(vect_dist) + s(std_", j, ") + s(X,Y)")
}

@@

summary(gam4$gam) ## summary of gam

plot(gam4$gam,pages=1)
```

```{r,  results="hide", include=FALSE}
plot_parametric(gam4$gam, pred=list(time=c("begin","end"), cell_type=c("control", "source")), parametricOnly = TRUE, main="", xlab ="log_density"
  #, groups = cell_type, gcolor = my_cols, color = my_cols["time"]
  ) 
```

## Immatures
```{r}
idata <- vocc_by_sp[["imm"]]
#View(data)
```

### Plot matches
```{r}
ggplot(data, 
#ggplot(filter(data, as.numeric(matchobs) <= 550), 
  aes(y=Y, x=X, shape=as.factor(treat), colour=matchobs)) +
  geom_point(alpha=0.4, size = 2) +
  scale_shape_manual(values= c(0, 15)) +
  scale_colour_viridis_d(option="C", guide = FALSE) +
  theme_bw()

@@

### Basic BACI
```{r}
mi <- lmerTest::lmer(as.formula(baci_formula), data=idata)
summary(mi)
```

```{r}
sjPlot::plot_model(mi, type = "std2", order.terms=c(1,2,8,3,4,5,6,7), title = paste(species, "immature"))
```

### GAMM
```{r}
# for (j in (variable_names)) {
#   gamm_formula <- paste0("log_density ~ time * cell_type + s(vect_dist) + s(std_", j, ") + s(X,Y)")
# }

@@

summary(igam4$gam) ## summary of gam

plot(igam4$gam,pages=1)
```

```{r, results="hide", include=FALSE}
plot_parametric(igam4$gam, pred=list(time=c("begin","end"), cell_type=c("control", "source")), parametricOnly = TRUE, main="", xlab ="log_density"
  #, groups = cell_type, gcolor = my_cols, color = my_cols["time"]
  ) 
```

Misc exploring gamm 
```{r eval=FALSE}
# summary(igam4$mer) ## underlying mixed model
# anova(igam4$gam)

# itsadug::plot_parametric(gam4$gam, pred=list(cell_type=c("control", "source"), after=c("0","1")), parametricOnly = TRUE, main="")

# mi <- mgcv::gamm(log_density~ as.factor(after) + cell_type + after:cell_type + std_temp + std_temp2 + vect_dist + vect_dist2 + s(as.factor(idata$icell), bs="re") + s(as.factor(idata$matchobs), bs="re"), data=idata)
# summary(mi)
# 
# mi <- mgcv::gamm(log_density~ as.factor(after) * cell_type + s(std_temp) + s(vect_dist) + s(as.factor(idata$matchobs), bs="re") , data=idata) # + s(X,Y, bs="re")
# summary(mi)
# mi
# 
# 
# idata$after <- as.factor(idata$after)
# idata$matchobs <- as.factor(idata$matchobs)
# 
# m4b <- gamm4::gamm4(log_density~ after * cell_type + s(std_temp) + s(vect_dist) + #s(X) + s(Y) + 
#     s(X,Y), random = ~(1|matchobs), data=idata) # + s(X,Y, bs="re")
# 
# 
# plot(m4b$gam,pages=1)
# 
# summary(m4b$gam) ## summary of gam
# summary(m4b$mer) ## underlying mixed model
# anova(m4b$gam) 
# 
# gamtabs(m1)
# 
# #library(itsadug)
# plot_parametric(m4b$gam, pred=list(cell_type=c("control", "source"), after=c("0","1")))
# 
# library(mgcv)
```
